Supplementary Materials And Methods

### 2019-09-17 Supplementary_mat_met.Rmd
article

---
title: "Supplementary File 1"
author: "Selcen Ari"
date: "7 Nisan 2019"
output:
  pdf_document: default
  html_document: default
  word_document: default
bibliography: ref.bib
---

#1. Defined functions for ceRNA models and workflow of method
We defined the functions that can be used with R programming. Briefly, these functions process a given miRNA:gene dataset and convert to graph object. All values that are significant in miRNA:target interactions are stored in edge variables and processed with formulations that are given in previous section. The functions and steps of approach are explained as following (Figure 1) :

@@

@@

|competing| miRNA | Competing_expression | miRNA_expression | seed_type | region | energy
|-----------|------------|----------|----------------|---------|-----------|----------|
| `Gene1`   | `Mir1` | 10000 | 1000 | 0.43 | 0.30 | -20 |
| `Gene2`   | `Mir1` | 10000 | 1000 | 0.43 | 0.01 | -15 |
| `Gene3`   | `Mir1` | 5000  | 1000 | 0.32 | 0.40 | -14 |
| `Gene4`   | `Mir1` | 10000 | 1000 | 0.23 | 0.50 | -10 |
| `Gene4`   | `Mir2` | 10000 | 2000 | 0.35 | 0.90 | -12 |
| `Gene5`   | `Mir2` | 5000  | 2000 | 0.05 | 0.40 | -11 |
| `Gene6`   | `Mir2` | 10000 | 2000 | 0.01 | 0.80 | -25 |

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", title = "Minimal dataset in steady-state conditions")
```

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  update_how("Gene2", 2)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation without interaction factors")
```

@@

@@

```
Note that the regulations are colored according to expression changes of present and a previous value. So, it can be observed that whole gene expressions increase in comparison of initial steady-state. The overall regulations of gene expressions are as followings:

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", title = "Minimal dataset in steady-state conditions")

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene2", 2)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation without interaction factors")
```

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene2", 2)%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation without interaction factors",2)
```

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene4", 2)%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation without interaction factors",2)
```

@@

#2. Obtaining breast cancer dataset and integration

@@

```{r, echo=FALSE, message= FALSE, warning=FALSE}
# Install the packages which is necessary to obtain and process datasets.
install.packages("BiocManager")
BiocManager::install("SummarizedExperiment")
BiocManager::install("TCGAbiolinks")
library(TCGAbiolinks)
library(SummarizedExperiment)
#install devtools
devtools::install_github("selcenari/ceRNAnetsim")
library(ceRNAnetsim)
```

@@

```{r,message= FALSE, warning=FALSE}

@@

```{r, message= FALSE, warning=FALSE}
BCP_normal <- GDCquery(project = "TCGA-BRCA",
                  data.category = "Transcriptome Profiling",
                  data.type = "Gene Expression Quantification", 
                  workflow.type = "HTSeq - Counts",
                  barcode = "TCGA-E9-A1N5-11A-41R-A14D-07")
GDCdownload(BCP_normal)

@@

```{r, echo=FALSE, message= FALSE, warning=FALSE}
#remove:
rm(BCPGE_normal, BCPGE_tumor, BCP_normal, BCP_tumor)
```

@@

```{r, message= FALSE, warning=FALSE}
BCP_mirnatumor <- GDCquery(project = "TCGA-BRCA",
                  data.category = "Transcriptome Profiling",
                  data.type = "Isoform Expression Quantification",
                  workflow.type = "BCGSC miRNA Profiling",
                  barcode = "TCGA-E9-A1N5-01A-11R-A14C-13")

@@

```{r, message= FALSE, warning=FALSE}
BCP_mirnanormal <- GDCquery(project = "TCGA-BRCA",
                  data.category = "Transcriptome Profiling",
                  data.type = "Isoform Expression Quantification",
                  workflow.type = "BCGSC miRNA Profiling",
                  barcode = "TCGA-E9-A1N5-11A-41R-A14C-13")

@@

+ The datasets were directly downloaded from supplementary data files of the studies.
+ It was provided that the datasets are converted to same human genome build.
+ The seed type information was organized as the same style.
+ The datasets were combined.
+ We commited the interaction factors as numeric values according to previous studies. (We added the interaction factors and their numeric values at following tables)

Note: Energy values in miRNA:target pairs are represented by high-throughput studies [@helwak_mapping_2013; @moore_mirnatarget_2015] which are utilized in this study. On the other hand, we have specified the other interaction factors, seed type and location of binding region on the target, as numeric values based on the previous studies.[@grimson_microrna_2007] have compared the seed types' effect on target repression with few miRNA had canonical seed pairing in their study. Additionally, [@bartel_micrornas:_2009] and [@betel2010comprehensive] have studied on functional and non-functional seed interactions. Based on results of these studies we have arranged seed types of miRNA:target interactions as numeric values. We also have redefined location of binding region on the target as numeric values, based on studies of [@hausser_analysis_2013] and [@helwak_mapping_2013]. With this process, we have handled this entegrated dataset in context of competitor behaviours and functionality of interactions.

|seed_type| Coefficient |
|-----------|------------|
| `5-mer`   | `0.05` |
| `5-mer_noncanonical`   | `0.04` |
| `6-mer`   | `0.07` |
| `6-mer_noncanonical`   | `0.05` |
| `6-merA1`   | `0.07` |
| `6-merA1_noncanonical`   | `0.05` |
| `7-mer`   | `0.23` |
| `7-mer_noncanonical`   | `0.19` |
| `7-merA1`   | `0.19` |
| `7-merA1_noncanonical`   | `0.16` |
| `7-mer-8m`   | `0.25` |
| `7-mer-8m_nonacnonical`   | `0.21` |
| `8-mer`   | `0.43` |
| `8-mer_noncanonical`   | `0.35` |
| `9-mer`   | `0.43` |
| `9-mer_noncanonical`   | `0.35` |
| `none`   | `0.01` |

|location of binding region on target| Coefficient |
|-----------|------------|
| `3'UTR`   | `0.84` |
| `CDS`   | `0.42` |
| `5'UTR`   | `0.01` |

@@

```{r, message= FALSE, warning=FALSE}
experimentalmirnagene%>%
  dplyr::select(miRNA, Ensembl_Gene_Id, hgnc_symbol, Energy, seed_type_effect, region_effect)%>%
  distinct()%>%
  group_by(Ensembl_Gene_Id, miRNA)%>%
  mutate(seed_type_effect= ifelse(seed_type_effect==max(seed_type_effect), seed_type_effect, max(seed_type_effect)), Energy = ifelse(Energy==min(Energy), Energy, min(Energy)), region_effect= ifelse(region_effect==max(region_effect), region_effect, max(region_effect)))%>%
  distinct()-> tocombine_mirnagene

@@

```{r, message= FALSE, warning=FALSE}
BCPME_mirnanormal%>%
  dplyr::inner_join(tocombine_mirnagene, by = c("miRNA_name"="miRNA"))%>%
  dplyr::inner_join(TCGA_E9_A1N5_normal, by = c("Ensembl_Gene_Id"="ensembl_gene_id", "hgnc_symbol"="external_gene_name"))%>%
  distinct()%>%
  dplyr::select(hgnc_symbol, miRNA_name, mirna_RPM= reads_per_million_miRNA_mapped, GE_normal, Energy, seed_type_effect, region_effect)-> E9GE_mirnagenenormal

head(E9GE_mirnagenenormal)
```

```{r, message= FALSE, warning=FALSE}
BCPME_mirnatumor%>%
  dplyr::inner_join(tocombine_mirnagene, by = c("miRNA_name"="miRNA"))%>%
  dplyr::inner_join(TCGA_E9_A1N5_tumor, by = c("Ensembl_Gene_Id"="ensembl_gene_id", "hgnc_symbol"="external_gene_name"))%>%
  distinct()%>%
  dplyr::select(hgnc_symbol, miRNA_name, mirna_RPM= reads_per_million_miRNA_mapped, GE_tumor, Energy, seed_type_effect, region_effect)-> E9GE_mirnagenetumor

head(E9GE_mirnagenetumor)

@@

@@

```{r}
E9GE_mirnagenenormal%>%
  priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
  find_node_perturbation(how = 3, cycle = 10, fast = 5)%>%
  filter(!is.na(perturbation_efficiency))->perturbationofnetwork

@@

```{r}
three_fold_change%>%
  inner_join(perturbationofnetwork, by = c("hgnc_symbol"="name"))

@@

```{r}
as.data.frame(E9GE_mirnagenenormal)%>%
 priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
 update_how("SERPINE2",2.75) %>%
 simulate(250) %>%
 find_iteration(limit=1, plot= TRUE) #limit=1 describes the change that is not taken into account.
```

@@

###2.5.3 Comparison of simulation results and tumor tissue expression values
```{r}
E9GE_mirnagenetumor%>%
  dplyr::select(hgnc_symbol, GE_tumor)%>%
  dplyr::inner_join((E9GE_mirnagenenormal%>%dplyr::select(hgnc_symbol, GE_normal)), by ="hgnc_symbol")%>%
  inner_join(simulation_results, by= c("hgnc_symbol"="name"))%>%
  distinct()
```

@@

>>>>

---
title: "Supplementary Materials And Methods"
author: "Selcen Ari"
date: "7 Nisan 2019"
output:
  pdf_document: default
  html_document: default
  word_document: default
bibliography: ref.bib
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0}     \renewcommand{\thefigure}{S\arabic{figure}}}
  \usepackage{subfig}
---

\beginsupplement

# 1. Defined functions for ceRNA models and workflow of method
We defined the functions that can be used with R programming. Briefly, these functions process a given miRNA:gene dataset and convert to graph object. All values that are significant in miRNA:target interactions are stored in edge variables and processed with formulations that are given in previous section. The functions and steps of approach are explained as following (Figure S1) :

@@

```{r, include=FALSE, message=FALSE, warning=FALSE}

library(tidygraph)
library(dplyr)
library(tidyr)
library(stringr)

@@

See Figure S1 in Supplementary Tables file.

@@

```{r, message=FALSE, warning=FALSE, fig.cap= "Minimal Dataset in Steady-state", fig.width=5, fig.height=4}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", title = "Minimal dataset in steady-state conditions")
```

@@

```{r, message=FALSE, warning=FALSE, fig.keep='all', fig.cap= "Gene2 Upregulation on Minimal Dataset", fig.width=5, fig.height=4}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  update_how("Gene2", 2)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation without interaction factors")
```

@@

```{r, message=FALSE, warning=FALSE, out.width='50%', fig.asp=1, fig.ncol = 2, fig.cap = "Sequential iteration of minsampdata", fig.subcap= c("First response of system to Gene2 upregulation (2nd iteration)", "Spreading of perturbation on system (3th iteration)")}

@@

Note that the regulations are colored according to expression changes of present and a previous value. So, it can be observed that whole gene expressions increase in comparison of initial steady-state. The overall regulations of gene expressions are as followings:

@@

```{r, message=FALSE, warning=FALSE, fig.cap= "Minimal Dataset with interaction factors in Steady-state", fig.width=5, fig.height=4}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", title = "Minimal dataset with interaction factors in steady-state conditions")

@@

```{r, message=FALSE, warning=FALSE, fig.cap= " When Gene2 is upregulated on Minimal Dataset with interaction factors", fig.width=5, fig.height=4}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene2", 2)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation with interaction factors")
```

```{r, message=FALSE, warning=FALSE, out.width='50%', fig.asp=1, fig.ncol = 2, fig.cap = "Sequential iteration of minsampdata with interaction factors", fig.subcap= c("First response of system to Gene2 upregulation (2nd iteration)", "Spreading of perturbation on system (3th iteration)")}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene2", 2)%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation with interaction factors",2)
```

@@

```{r, message=FALSE, warning=FALSE, out.width='50%', fig.asp=1, fig.ncol = 2, fig.cap = "Perturbation of Gene4 on minsampdata with interaction factors", fig.subcap= c("1st iteration after Gene4 upregulation", "2nd iteration after Gene4 perturbation")}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene4", 2)%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Common Gene-Gene4 Upregulation without interaction factors",2)
```

@@

# 2. Obtaining breast cancer dataset and integration

@@

```{r, echo=FALSE, message= FALSE, warning=FALSE}
# Install the packages which is necessary to obtain and process datasets.
#install.packages("BiocManager")
#BiocManager::install("SummarizedExperiment")
#BiocManager::install("TCGAbiolinks")
library(TCGAbiolinks)
library(SummarizedExperiment)
```

@@

```{r,message= FALSE, warning=FALSE, eval=FALSE}

@@

```{r, message= FALSE, warning=FALSE, eval=FALSE}
BCP_normal <- GDCquery(project = "TCGA-BRCA",
                  data.category = "Transcriptome Profiling",
                  data.type = "Gene Expression Quantification", 
                  workflow.type = "HTSeq - Counts",
                  barcode = "TCGA-E9-A1N5-11A-41R-A14D-07")
GDCdownload(BCP_normal)

@@

@@

```{r, message= FALSE, warning=FALSE, eval=FALSE}
BCP_mirnatumor <- GDCquery(project = "TCGA-BRCA",
                  data.category = "Transcriptome Profiling",
                  data.type = "Isoform Expression Quantification",
                  workflow.type = "BCGSC miRNA Profiling",
                  barcode = "TCGA-E9-A1N5-01A-11R-A14C-13")

@@

```{r, message= FALSE, warning=FALSE, eval=FALSE}
BCP_mirnanormal <- GDCquery(project = "TCGA-BRCA",
                  data.category = "Transcriptome Profiling",
                  data.type = "Isoform Expression Quantification",
                  workflow.type = "BCGSC miRNA Profiling",
                  barcode = "TCGA-E9-A1N5-11A-41R-A14C-13")

@@

+ The datasets were directly downloaded from supplementary data files of the studies.
+ It was provided that the datasets are converted to same human genome build.
+ The seed type information was organized as the same style.
+ The datasets were combined.
+ We commited the interaction factors as numeric values according to previous studies. (We added the interaction factors and their numeric values at Supplementary tables S2-3)

@@

```{r, message= FALSE, warning=FALSE, eval=FALSE}
experimentalmirnagene%>%
  dplyr::select(miRNA, Ensembl_Gene_Id, hgnc_symbol, Energy, seed_type_effect, region_effect)%>%
  distinct()%>%
  group_by(Ensembl_Gene_Id, miRNA)%>%
  mutate(seed_type_effect= ifelse(seed_type_effect==max(seed_type_effect), seed_type_effect, max(seed_type_effect)), Energy = ifelse(Energy==min(Energy), Energy, min(Energy)), region_effect= ifelse(region_effect==max(region_effect), region_effect, max(region_effect)))%>%
  distinct()-> tocombine_mirnagene

@@

```{r, message= FALSE, warning=FALSE, eval=FALSE}
BCPME_mirnanormal%>%
  dplyr::inner_join(tocombine_mirnagene, by = c("miRNA_name"="miRNA"))%>%
  dplyr::inner_join(TCGA_E9_A1N5_normal, by = c("Ensembl_Gene_Id"="ensembl_gene_id", "hgnc_symbol"="external_gene_name"))%>%
  distinct()%>%
  dplyr::select(hgnc_symbol, miRNA_name, mirna_RPM= reads_per_million_miRNA_mapped, GE_normal, Energy, seed_type_effect, region_effect)-> E9GE_mirnagenenormal

@@

```{r}
data("E9GE_mirnagenenormal")
head(E9GE_mirnagenenormal)
```

```{r, message= FALSE, warning=FALSE, eval=FALSE}
BCPME_mirnatumor%>%
  dplyr::inner_join(tocombine_mirnagene, by = c("miRNA_name"="miRNA"))%>%
  dplyr::inner_join(TCGA_E9_A1N5_tumor, by = c("Ensembl_Gene_Id"="ensembl_gene_id", "hgnc_symbol"="external_gene_name"))%>%
  distinct()%>%
  dplyr::select(hgnc_symbol, miRNA_name, mirna_RPM= reads_per_million_miRNA_mapped, GE_tumor, Energy, seed_type_effect, region_effect)-> E9GE_mirnagenetumor

```{r}
data("E9GE_mirnagenetumor")
head(E9GE_mirnagenetumor)
```

@@

```{r, include=FALSE, eval=FALSE}
as.data.frame(E9GE_mirnagenenormal)%>%
  priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
  find_node_perturbation(how = 3, cycle = 10, fast = 5)%>%
  filter(!is.na(perturbation_efficiency))->perturbationofnetwork

@@

```{r, eval=FALSE}
three_fold_change%>%
  inner_join(perturbationofnetwork, by = c("hgnc_symbol"="name"))

@@

```{r, fig.cap= "Percentage of affected nodes of each iteration for SERPINE2 Gene"}
as.data.frame(E9GE_mirnagenenormal)%>%
 priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
 update_how("SERPINE2",2.75) %>%
 simulate(150) %>%
 find_iteration(limit=1, plot= TRUE) #limit=1 describes the change that is not taken into account.
```

@@

### 2.5.3 Comparison of simulation results and tumor tissue expression values
```{r}
E9GE_mirnagenetumor%>%
  dplyr::select(hgnc_symbol, GE_tumor)%>%
  dplyr::inner_join((E9GE_mirnagenenormal%>%dplyr::select(hgnc_symbol, GE_normal)), by ="hgnc_symbol")%>%
  inner_join(simulation_results, by= c("hgnc_symbol"="name"))%>%
  distinct()
```

@@

# 3. REFERENCES
